Mutual Information Plots

### 2025-11-01 mutual_information_curves.qmd
Fixed combinations issue in weighted Kendall's

weighted_kendall <- function(rhat, k, wtype = "symmetric_salience", salient_weights = NULL) {
  n <- length(rhat)
  #salient_weights needs to be a vector of length k

  if(is.null(salient_weights)){
    salient_weights <- rep(1,k)
  }

  if (is.null(wtype)) {
    w <- matrix(0, ncol = n, nrow = n)
    for (q in k:1){
      w[q, ] <- salient_weights[q]
      w[, q] <- salient_weights[q]
    }
    w <- w/2
  }

  if (wtype == "symmetric_salience") {
    w <- matrix(0, ncol = n, nrow = n)
    for (q in k:1) {
      w[q, ] <- salient_weights[q]
      w[, q] <- salient_weights[q]
    }
    #w[upper.tri(w)] <- w[upper.tri(w)]/2
    #w[lower.tri(w)] <- w[lower.tri(w)]/2
  }

  #Ryan method
  if (wtype == "rydog"){
  w <- matrix(0, ncol = n, nrow = n)
#  w[1:k, 1:k] <- 1
  for (q in k:1) {
    w[1:q, 1:q] <- salient_weights[q]
    w[1:q, 1:q] <- salient_weights[q]
  }
  #w[upper.tri(w)] <- w[upper.tri(w)]/2
  #w[lower.tri(w)] <- w[lower.tri(w)]/2
  }

  summ <- 0
  for (i in 1:n) {
    for (j in 1:n) {
      if (i != j) {
        summ <- summ + w[i, j] * sign(i - j) * sign(rhat[i] - rhat[j])
        #print(summ)
      }

    }
  }

  out <- summ / (sum(w) - sum(diag(w)))
  weight <- (out + 1) / 2
  #print(sum(w))
  #print(sum(diag(w)))
  return(weight)

}
weighted_kendall(c(2,1,3,4), k=4, salient_weights = c(3,1,1,0))
weighted_kendall(c(2,1,3,4), k=2)

@@

>>>>

weighted_kendall <- function(rhat, k, wtype = "symmetric_salience", salient_weights = NULL, r = NULL) {
  n <- length(rhat)
  if (is.null(r)) r <- 1:n
  if (is.null(salient_weights)) salient_weights <- rep(1, k)

  # Weight matrix
  w <- matrix(0, ncol = n, nrow = n)
  if (is.null(wtype) || wtype == "symmetric_salience") {
    for (q in k:1) {
      w[q, ] <- salient_weights[q]
      w[, q] <- salient_weights[q]
    }
  }
  if (wtype == "rydog") {
    for (q in k:1) {
      w[1:q, 1:q] <- salient_weights[q]
    }
  }

  # Compute weighted Kendall numerator and denominator consistently
  summ <- 0
  denom <- 0
  for (i in 1:n) {
    for (j in 1:n) {
      if (i != j && w[i, j] != 0) {
        # Skip uninformative pairs (ties in either vector)
        if (r[i] == r[j] || rhat[i] == rhat[j]) next

        denom <- denom + abs(w[i, j])
        summ <- summ + w[i, j] * sign(r[i] - r[j]) * sign(rhat[i] - rhat[j])
      }
    }
  }

  # Handle case where denom = 0 (e.g., all ties)
  if (denom == 0) return(1)

  out <- summ / denom
  weight <- (out + 1) / 2
  return(weight)
}

weighted_kendall(c(2,1,3,4), k=4, salient_weights = c(3,1,1,0))
weighted_kendall(c(2,1,3,4), k=2)

weighted_kendall(c(1,1,1,1,5,6), k = 6, r = c(1,1,1,1,5,6))
# Only works for k = # of top teams tied

@@

## Combinations

-   Manually setting the top-K teams with r=1 to look at if the top K teams were in the top K spots

```{r}
compute_mut_plots_combinations <- function(dataset, num_teams) {

  results <- numeric(num_teams)

  for (k in 1:num_teams) {
    df <- dataset %>%
      mutate(
        true_rank = ifelse(true_rank <= k, 1, true_rank),
        rank_hat = ifelse(rank_hat <= k, 1, rank_hat)
      )
    binary_vector <- c(rep(1, k), rep(0, num_teams - k))
    results[k] <- mut_info(df, num_teams, binary_vector)
  }
  return(results)
}

combinations_se <- compute_mut_plots_combinations(all_results_norm, num_teams=8)
combinations_rr <- compute_mut_plots_combinations(test_rr_noTies, num_teams=8)

# Define x values
x <- 1:8

# Combine into one data frame for plotting
plot_data <- data.frame(
  x = rep(x, 2),
  y = c(1-mi_se_norm, 1-combinations_se, 1-mi_test_rr, 1-combinations_rr),
  type = rep(c("mi_se_norm", "combinations_se", "mi_test_rr", "combinatinos_rr"), each = 8)
)

# Create plot
ggplot(plot_data, aes(x = x, y = y, color = type, group = type)) +
  geom_line(size = 1.2) +
  geom_point(size = 3) +
  scale_x_continuous(breaks = 1:8) +
  labs(
    x = "Salient Ranks",
    y = "Value",
    color = "Series",
    title = "Comparison of MI vs Top K-teams"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5)
  )
